Efficient Calculation of Interaction Features on Sparse Matrices

### 2015-10-07 paper/spares_interaction.tex
fixed an index in the algo

\section{Approach}
Let the list of nonzero columns for a given row $\vec{x}$ be denoted by $N_{zc}$. The nonzero second degree 
interaction features are simply the products of all combinations of two elements whose 
columns are in $N_{zc}$. However, to properly place an interaction feature into the correct column, a mapping from the column 
index pairs of $N_{zc}$ into the columns of the interaction matrix is needed. The mapping is 
from the space (a, b) where a,b are in $1,2,..., D$ onto the space $1,2,..., \frac{D^2-D}{2}$. This 
is isomorphic to mapping the coordinates of the upper triangle of a matrix onto a flat 
list. The following is a proof by construction for such a mapping.

@@

\begin{codebox}
\Procname{$\proc{Sparse Interaction}(A)$}
    \zi $\func{map}(a, b) = \frac{2Da-a^2+2b-3a-2}{2}$
    \zi $N \gets$ row count of $A$
    \zi $D \gets$ column count of $A$
    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
    \zi \For $\id{row}$ in $A$ \Do
    \zi     $N_{zc} \gets$ nonzero columns of $row$
    \zi     \For $i \gets 0 \To |N_{zc}|-2$ \Do
    \zi         \For $j \gets i+1 \To |N_{zc}|-1$ \Do
    \zi             $k \gets \func{map}(i, j)$
    \zi             $r \gets$ index of $\id{row}$
    \zi             $B[r, k] \gets \id{row}[i] \cdot \id{row}[j]$
                \End
            \End
       	\End
\end{codebox}

>>>>

\section{Approach}
Let the list of nonzero columns for a given row $\vec{x}$ be denoted by $N_{zc}$. The nonzero second degree 
interaction features are simply the products of all combinations of two elements whose 
columns are in $N_{zc}$. However, to properly place an interaction feature into the correct column, a mapping from the column 
index pairs of $N_{zc}$ into the columns of the interaction matrix is needed. The mapping is 
from pairs $(a, b)$,  where $a$ and $b$ are in $1,2,..., D$, and $a < b$, to $1,2,..., \frac{D^2-D}{2}$,  Such a mapping essentially consists of mapping the indices of entries in the upper triangle of a matrix to indices in a flat 
list. We now describe the construction of such a mapping.

\subsection{Mapping construction}
We seek a map from matrix indices $(i, j)$ (with $i < j$ and $0 \le i < D$) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
x & 0 & 1 & 2 \\
x & x & 3 & 4 \\
x & x & x & 5 \\
x & x & x & x
\end{bmatrix}
\end{align}
It's considerably easier, however, to consider the same indices, but subtracted from $6$ (or more generally, from $\frac{D(D-1)}{2}$; that gives the pattern
\begin{align}
\label{eq:4x4mat}
\begin{bmatrix}
x & 6 & 5 & 4 \\
x & x & 3 & 2 \\
x & x & x & 1 \\
x & x & x & x
\end{bmatrix}
\end{align}
We'll call the function defined by this example $(i, j) \mapsto g(i, j)$, and then observe that 
\begin{align}
f(i, j) = \frac{D(D-1)}{2} - g(i, j)
\end{align}

To simplify slightly, we introduce a notation for the $n$th triangular number, 
\begin{align}
\begin{bmatrix}
T_2(n) = \frac{n(n+1)}{2}
\end{bmatrix}
\end{align}
\noindent
The subscript $2$ is there to indicate that these are triangles in two dimensions; we'll use $T_3(n)$ to indicate the $n$th tetrahedral number, and so on for higher dimensions.

The codomain of $g$ is now numbers from $1$ to $T_2(D-1)$, inclusive.

Observe that in Equation~\ref{eq:4x4mat}, each entry in row $i$ lies in the range
\begin{align}
T_2(D-i-1) &< e \le T_2(D-i).
\end{align}
\noindent
For instance, in row $2$ in our example, where $D = 4$, the entries range from $2$ to $3$, while $T_2(D-i-1) = T_2(1) = 1$ and $T_2(D-i) = T_2(2) = 3$. (Note that row indices start at zero.) Unfortunately, the numbers increase from right to left. The entry in column $j$ is just $T_2(D-i-1) + D-j$, which adds one for the rightmost column (because $D - (D-1) = 1$). 
Thus, the formula for $g$ is simply
\begin{align}
g(i, j) 
&= T_2(D-i-1) + (D-j) \\
&= \frac{(D-i-1)(D-i)}{2} + D-j \\
&= \frac{(D^2 - (2i)D - D - i^2 - i) + 2D - 2j}{2} \\
&= \frac{D^2 - (2i)D + D - i^2 - i - 2j}{2}
\end{align}
and hence 
\begin{align}
f(i, j) 
&= \frac{D(D-1)}{2} - g(i,j) \\
&= \frac{D^2-D}{2} - \frac{D^2 - (2i)D + D - i^2 - i - 2j}{2}\\
&= \frac{D^2-D - D^2 + (2i)D - D + i^2 + i + 2j}{2}\\
&= \frac{(2i)D - 2D + i^2 + i + 2j}{2}
\end{align}
[WRONG]
Correct result:
\begin{align}
R = \frac{ 2in - i^2  + 2j - 3i - 2}{2}.
\end{align}
\subsubsection{Other indices}
With one-based indexing, the formula above becomes
\begin{align}
f_1(i, j) &= ....
\end{align}

  column

@@

\begin{codebox}
\Procname{$\proc{Sparse Interaction}(A)$}
    \zi $\func{map}(a, b) = \frac{2Da-a^2+2b-3a-2}{2}$
    \zi $N \gets$ row count of $A$
    \zi $D \gets$ column count of $A$
    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
    \zi \For $\id{row}$ in $A$ \Do
    \zi     $N_{zc} \gets$ nonzero columns of $row$
    \zi     \For $i \gets 0 \To |N_{zc}|-1$ \Do
    \zi         \For $j \gets i+1 \To |N_{zc}|$ \Do
    \zi             $k \gets \func{map}(i, j)$
    \zi             $r \gets$ index of $\id{row}$
    \zi             $B[r, k] \gets \id{row}[i] \cdot \id{row}[j]$
                \End
            \End
       	\End
\end{codebox}
